---
title: "Data Validation Report"
subtitle: "Cross-checking africa_measles.csv against WHO and World Bank source APIs"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
    output-file: data-validation-report.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
library(dplyr)
library(scales)
```

## Overview

This report validates the dataset used in the [Africa Measles Heatmap](https://bk-advisors.github.io/africa-measles/) against its two original sources:

| Column | Source | API |
|--------|--------|-----|
| `cases` | WHO Global Health Observatory | Indicator **WHS3_62** --- Measles reported cases |
| `population` | World Bank Open Data | Indicator **SP.POP.TOTL** --- Population, total |
| `cases_per_million` | Derived | `cases / population * 1,000,000` |

Data was pre-fetched from both APIs using `R/fetch_validation_data.R` and cached locally. To refresh, re-run that script before rendering this report.

```{r load-data}
csv   <- read.csv("../data/africa_measles.csv")
cache <- readRDS("../data/validation_cache.rds")

cat("CSV rows:", nrow(csv), "\n")
cat("Countries:", n_distinct(csv$country), "\n")
cat("Year range:", min(csv$year), "-", max(csv$year), "\n")
cat("API data fetched:", format(cache$fetched_at), "\n")
```

## 1. Measles cases --- WHO GHO API

### The indicator issue

The WHO GHO API hosts **six** measles-related indicators. During initial validation, indicator `WHS3_41` was used based on third-party documentation. This produced alarming discrepancies --- only 161 of 1,226 joined rows matched, and those were almost entirely zero-valued cells.

Investigation revealed the problem: `WHS3_41` is a different measure entirely. Here is how the two indicators compare for a single data point:

| Indicator | Description | Kenya 1983 value |
|-----------|-------------|------------------|
| `WHS3_41` | *(not reported case counts)* | **`r cache$wrong_sample$NumericValue[1]`** |
| **`WHS3_62`** | **Measles --- number of reported cases** | **`r comma(cache$right_sample$NumericValue[1])`** |

The CSV value for Kenya 1983 is **`r comma(csv$cases[csv$country_code == "KEN" & csv$year == 1983])`** --- matching `WHS3_62` exactly.

### All WHO measles indicators

```{r who-indicators}
cache$indicators %>%
  knitr::kable(col.names = c("Code", "Indicator Name"))
```

### Comparison: CSV vs WHO (WHS3_62)

```{r compare-cases}
cases_check <- csv %>%
  select(country, country_code, year, csv_cases = cases) %>%
  inner_join(cache$who, by = c("country_code", "year")) %>%
  mutate(diff = csv_cases - who_cases)

exact <- sum(cases_check$diff == 0)
total <- nrow(cases_check)

cat("WHO records available:", nrow(cache$who), "\n")
cat("Joined rows:", total, "of", nrow(csv), "CSV rows\n")
cat("Exact matches:", exact, "/", total, "\n")
```

```{r cases-verdict, results='asis'}
if (exact == total) {
  cat("::: {.callout-tip}\n## PASS\nEvery case count in the CSV matches the WHO source exactly.\n:::\n")
} else {
  mismatches <- cases_check %>% filter(diff != 0)
  cat("::: {.callout-warning}\n## DISCREPANCIES\n", nrow(mismatches), "rows differ.\n:::\n")
  mismatches %>%
    select(country, year, csv_cases, who_cases, diff) %>%
    head(20) %>%
    knitr::kable()
}
```

## 2. Population --- World Bank API

```{r compare-pop}
pop_check <- csv %>%
  select(country, country_code, year, csv_pop = population) %>%
  inner_join(cache$wb, by = c("country_code", "year")) %>%
  mutate(diff = csv_pop - wb_pop)

exact_p <- sum(pop_check$diff == 0)
total_p <- nrow(pop_check)

cat("World Bank records available:", nrow(cache$wb), "\n")
cat("Joined rows:", total_p, "of", nrow(csv), "CSV rows\n")
cat("Exact matches:", exact_p, "/", total_p, "\n")
```

```{r pop-verdict, results='asis'}
if (exact_p == total_p) {
  cat("::: {.callout-tip}\n## PASS\nEvery population figure in the CSV matches the World Bank source exactly.\n:::\n")
} else {
  pop_mis <- pop_check %>% filter(diff != 0)
  cat("::: {.callout-warning}\n## DISCREPANCIES\n", nrow(pop_mis), "rows differ.\n:::\n")
  pop_mis %>%
    select(country, year, csv_pop, wb_pop, diff) %>%
    head(20) %>%
    knitr::kable()
}
```

## 3. Derived column --- `cases_per_million`

This column should equal `cases / population * 1,000,000`, rounded to one decimal place.

```{r check-rate}
csv <- csv %>%
  mutate(
    calc_rate = round(cases / population * 1e6, 1),
    rate_diff = abs(cases_per_million - calc_rate)
  )

rate_ok <- sum(csv$rate_diff < 0.2)
cat("Consistent (within rounding tolerance of 0.2):", rate_ok, "/", nrow(csv), "\n")
```

```{r rate-verdict, results='asis'}
if (rate_ok == nrow(csv)) {
  cat("::: {.callout-tip}\n## PASS\nAll rates are internally consistent with cases and population.\n:::\n")
} else {
  cat("::: {.callout-warning}\n## ISSUES\n:::\n")
  csv %>% filter(rate_diff >= 0.2) %>%
    select(country, year, cases, population, cases_per_million, calc_rate, rate_diff) %>%
    head(10) %>%
    knitr::kable()
}
```

## 4. Coverage gaps

```{r coverage, results='asis'}
who_codes <- unique(cache$who$country_code)
csv_codes <- unique(csv$country_code)

in_who_not_csv <- setdiff(who_codes, csv_codes)

if (length(in_who_not_csv) > 0) {
  cat("African countries present in WHO data but excluded from the CSV:\n\n")
  cat(paste0("**", paste(in_who_not_csv, collapse = "**, **"), "**\n\n"))
  cat("The chart requires 15+ years of reporting; these were likely excluded for sparse data.\n")
} else {
  cat("All WHO African countries are represented in the CSV.\n")
}
```

## 5. Summary

```{r summary-table}
results <- data.frame(
  Check = c(
    "Measles cases vs WHO (WHS3_62)",
    "Population vs World Bank (SP.POP.TOTL)",
    "Derived rate (cases_per_million)",
    "All CSV rows joined to WHO",
    "All CSV rows joined to World Bank"
  ),
  Result = c(
    ifelse(exact == total, "PASS", "FAIL"),
    ifelse(exact_p == total_p, "PASS", "FAIL"),
    ifelse(rate_ok == nrow(csv), "PASS", "FAIL"),
    ifelse(total == nrow(csv), "PASS", paste0(total, " of ", nrow(csv))),
    ifelse(total_p == nrow(csv), "PASS", paste0(total_p, " of ", nrow(csv)))
  )
)

knitr::kable(results, align = c("l", "c"))
```

---

*Validation data fetched `r format(cache$fetched_at)` from the WHO GHO and World Bank APIs. Report rendered `r Sys.time()`.*
